---
title: 'KCNI Summer Academy 2023: Day 4'
author: "Mel Davie"
date: "13/07/2023"
output: pdf_document
---

# Case-control differential expression with pseudobulks
## Setup
### Set working directory
```{r setup}
knitr::opts_knit$set(root.dir = "~/Downloads/KCNISS_2023/SST_pseudobulk/")
```

### Load packages
```{r}
#install.packages("BiocManager")
#BiocManager::install("edgeR")
#BiocManager::install("locfit") 
##if locfit above doesn't work (because your version of R is too old):
#install.packages("https://cran.r-project.org/src/contrib/Archive/locfit/locfit_1.5-9.4.tar.gz", repos=NULL, type="source")
#install.packages("ggrepel")
#install.packages("cowplot")
#install.packages("gprofiler2")

suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(edgeR)
  library(ggrepel)
  library(cowplot)
  library(limma)
  library(ggpubr)
  library(gprofiler2)
})
```

### Load pseudobulked gene expression data
The data we're about to load in have been pseudobulked in advance using this script:
https://github.com/stripathy/pseudobulk_tutorial/blob/master/sea_ad_make_pseudobulk.R

We will only be working with one cell type: SST neurons. 

```{r}
pseudobulk_sst_counts = read_csv('SST_pseudobulk.csv')
sea_ad_meta = read_csv('SST_pseudobulk_meta.csv')
gene_info = read_csv('SST_pseudobulk_gene_info.csv')
rownames(pseudobulk_sst_counts) = gene_info$gene_symbol

sea_ad_meta = sea_ad_meta %>% mutate(disease = factor(disease, 
                                                      levels = c('normal', 'dementia')))
```

## Examining our dataset
How does the number of SST cells per person compare between controls and patients with dementia?

```{r}
sea_ad_meta %>% 
  ggplot(aes(x = disease, y = cells_per_donor)) + 
  geom_boxplot() + 
  theme_cowplot() + 
  ylab('SST cells per donor') +
  stat_compare_means()
```

It looks like there are fewer SST cells in individuals with dementia than controls

### Normalize our counts to CPM
```{r}
### Create a cpm object with normalized gene expression counts for each pseudobulk sample
pseudobulk_sst_cpm = edgeR::cpm(y = pseudobulk_sst_counts)
rownames(pseudobulk_sst_cpm) = gene_info$feature_name

pseudobulk_sst_cpm_trans = pseudobulk_sst_cpm %>% 
  t() %>% 
  as.data.frame()

pseudobulk_sst_cpm_trans = left_join(sea_ad_meta, pseudobulk_sst_cpm_trans %>% 
                                       rownames_to_column(var = 'donor_id')) 
```

### Plot gene expression of a gene (SST below) vs. all cell types
```{r}
pseudobulk_sst_cpm_trans %>% 
  ggplot(aes(x = disease, y = log2(SST+1))) + 
  geom_boxplot() + 
  theme_cowplot() + 
  ylab('SST expr (log2 CPM+1)') +
  stat_compare_means()
```

### Show how SST normalized expression is related to the number of cells sampled per donor
```{r}
pseudobulk_sst_cpm_trans %>% 
  ggplot(aes(x = log2(SST+1), y = log10(cells_per_donor + 1), color = disease)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  theme_cowplot() + 
  xlab('SST cells per donor (log10 units)') + 
  ylab('SST expr (log2 CPM+1)') +
  stat_cor()
```

## Differential expression with limma-voom
### Set up stats model
```{r}
# Find subjects with information for disease, Sex, PMI, and Age
use_subjects = complete.cases(sea_ad_meta %>% dplyr::select(disease, Sex, PMI, Age_norm))

# Drop subjects with fewer than 50 SST cells per donor
use_subjects = use_subjects & (sea_ad_meta$cells_per_donor > 50)

disease = factor(unlist(sea_ad_meta[use_subjects, 'disease']), 
                 levels = c('normal', 'dementia'))
sex = sea_ad_meta[use_subjects, 'Sex'] %>% unlist
pmi = sea_ad_meta[use_subjects, 'PMI'] %>% unlist
age = sea_ad_meta[use_subjects, 'Age_norm'] %>% unlist
cells_per_donor = sea_ad_meta[use_subjects, 'cells_per_donor'] %>% unlist

# Count the number of subjects for each disease group
sea_ad_meta[use_subjects, ] %>% 
  group_by(disease) %>% 
  tally
```

### Set up DGEList object
```{r}
dge0 = DGEList(pseudobulk_sst_counts[, use_subjects], genes = gene_info)

min_samples_expressing_gene <- sea_ad_meta[use_subjects, ] %>% nrow * 0.8 
# The 0.8 here refers to the fraction of total samples that needs to express the gene

dge0 = dge0[rowSums(dge0$counts >= 1) >= min_samples_expressing_gene, ] 
# This step filters genes such that they need to be detected in at least 80% of samples

# dge0 = DGEList(pb_counts, group = pb_metadata$Phenotype)
dge0 = calcNormFactors(dge0, method = "TMM")
```

### Set up design based on the factors defined above
```{r}
design = model.matrix(~ age + pmi + sex + log10(cells_per_donor) + disease) 

# Perform voom transformation
vm = voom(dge0, design, plot = TRUE)

# Perform lmFit and eBayes
fit = lmFit(vm, design)
fit = eBayes(fit)
```
From the plot above, can see that genes with low counts have already been filtered out (otherwise we'd see a downward dip in the trend towards the left—if you see this, you need to go back and filter).

### Analyze sex and disease terms
#### Analyze genes associated with sex
```{r}
deg_table_sex = topTable(fit, coef = "sexMale",  n = Inf, sort = "none", 
                         adjust.method = "BH")

deg_table_sex %>% 
  arrange(adj.P.Val) %>% 
  head()
```

#### Volcano plot of genes associated with sex
```{r}
sex_volcano = deg_table_sex %>% 
  ggplot(aes(x = logFC, y = -log10(P.Value), label = feature_name)) + 
  geom_point() +
  geom_text_repel(data = subset(deg_table_sex, adj.P.Val < 0.05), 
                  aes(label = feature_name), 
                  vjust = 1.5) +
  geom_point(data = subset(deg_table_sex, adj.P.Val < 0.05), 
             aes(color = 'red')) + 
  theme_cowplot() +
  theme(legend.position = "none")

sex_volcano
```

#### Analyze genes associated with disease
```{r}
deg_table_disease = topTable(fit, coef = "diseasedementia",  n = Inf, sort = "none", 
                             adjust.method = "BH")

deg_table_disease %>% 
  arrange(adj.P.Val) %>% 
  head(20)
```

#### Volcano plot of genes associated with disease
```{r}
disease_volcano = deg_table_disease %>% 
  ggplot(aes(x = logFC, y = -log10(P.Value), label = feature_name)) + 
  geom_point() +
  geom_text_repel(data = subset(deg_table_disease, adj.P.Val < 0.1), 
                  aes(label = feature_name), 
                  vjust = 1.5) + 
  geom_point(data = subset(deg_table_disease, adj.P.Val < 0.1), 
             aes(color = 'red')) + 
  theme_cowplot() +
  theme(legend.position = "none")

disease_volcano
```

#### Plot gene expression of a couple of DE genes below
```{r}
ihih5_plot = pseudobulk_sst_cpm_trans %>% 
  ggplot(aes(x = disease, y = log2(ITIH5+1))) + 
  geom_boxplot() + 
  theme_cowplot() +
  stat_compare_means()

ihih5_plot
```

```{r}
aqp4_as1_plot = pseudobulk_sst_cpm_trans %>% 
  ggplot(aes(x = disease, y = log2(`AQP4-AS1`+1))) + 
  geom_boxplot() + 
  theme_cowplot() +
  stat_compare_means()

aqp4_as1_plot
```

## Gene ontology analysis
### Disease
#### Get gene lists
```{r}
# Genes that have higher expression in individuals with dementia relative to controls
disease_up_ordered <- deg_table_disease %>% 
  filter(P.Value < 0.05 & logFC > 0) %>% 
  arrange(desc(logFC)) %>% 
  .$gene_symbol

# Genes that have lower expression in individuals with dementia relative to controls
disease_down_ordered <- deg_table_disease %>% 
  filter(P.Value < 0.05 & logFC < 0) %>% 
  arrange(logFC) %>% 
  .$gene_symbol
```

#### Enrichment analysis
```{r}
gp_disease = gost(list("up-regulated" = disease_up_ordered,
	  "down-regulated" = disease_down_ordered), organism = "hsapiens",
		        ordered_query = TRUE, sources = c("GO:BP", "GO:CC", "GO:MF"))

# Top terms
gp_disease_terms <- gp_disease$result %>% 
  group_by(query, source) %>% 
  slice_min(order_by = p_value, n = 3) %>% 
  .$term_id
```

#### Plot results
```{r}
gostplot(gp_disease, capped = FALSE, interactive = FALSE, 
         pal = list(`GO:MF` = "#6DC8E8", `GO:CC` = "#FF9179", `GO:BP` = "#57BCAC")) +
  geom_text_repel(mapping = aes(label = ifelse(gp_disease$result$term_id %in% 
                                                 gp_disease_terms,
                                               term_name, "")), 
                  force = 40, max.overlaps = Inf,size = 3) +
  geom_point(size = 3, colour = ifelse(gp_disease$result$term_id %in% 
                                                 gp_disease_terms,
                                       "black", "white"),
             alpha = ifelse(gp_disease$result$term_id %in% 
                                                 gp_disease_terms,
                                       1, 0))
```

### Sex
#### Get gene lists
```{r}
# Genes that have higher expression in males relative to females
male_up_ordered <- deg_table_sex %>% 
  filter(P.Value < 0.05 & logFC > 0) %>% 
  arrange(desc(logFC)) %>% 
  .$gene_symbol

# Genes that have lower expression in males relative to females
male_down_ordered <- deg_table_sex %>% 
  filter(P.Value < 0.05 & logFC < 0) %>% 
  arrange(logFC) %>% 
  .$gene_symbol
```

#### Enrichment analysis
```{r}
gp_sex = gost(list("up-regulated" = male_up_ordered,
	  "down-regulated" = male_down_ordered), organism = "hsapiens",
		        ordered_query = TRUE, sources = c("GO:BP", "GO:CC", "GO:MF"))

# Top terms
gp_sex_terms <- gp_sex$result %>% 
  group_by(query, source) %>% 
  slice_min(order_by = p_value, n = 3) %>% 
  .$term_id
```

#### Plot results
```{r}
gostplot(gp_sex, capped = FALSE, interactive = FALSE, 
         pal = list(`GO:MF` = "#6DC8E8", `GO:CC` = "#FF9179", `GO:BP` = "#57BCAC")) +
  geom_text_repel(mapping = aes(label = ifelse(gp_sex$result$term_id %in% 
                                                 gp_sex_terms,
                                               term_name, "")), 
                  force = 40, max.overlaps = Inf, size = 3) +
  geom_point(size = 3, colour = ifelse(gp_sex$result$term_id %in% 
                                                 gp_sex_terms,
                                       "black", "white"),
             alpha = ifelse(gp_sex$result$term_id %in% 
                                                 gp_sex_terms,
                                       1, 0))
```
