---
title: 'KCNI Summer Academy 2023: Day 3'
author: "Mel Davie"
date: "12/07/2023"
output: github_document
---

# Dataset integration & automated cell type annotation
## Setup
### Set working directory
```{r setup}
knitr::opts_knit$set(root.dir = "~/Downloads/KCNISS_2023/")
```

### Load packages
```{r}
library(tidyverse)
library(Seurat)

### might need the following:
# install.packages("BiocManager")
# BiocManager::install("multtest")
# library(multtest)
# 
# BiocManager::install("limma")
# library(limma)
# 
# install.packages("metap")
# library(metap)
```

### Create new Seurat objects
But first, let's make a couple changes:
- Just keep the genes that are present in both datasets
- Add a column to metadata specifying the source dataset for each sample

```{r}
# load smart-seq
meta_smart <- read.csv("AIBS_smart_meta_mini.csv", row.names = 1) %>% 
  mutate(method = "smart")
counts_smart <- read.csv("AIBS_smart_counts_mini.csv", row.names = 1)

# load 10x
meta_10x <- read.csv("SEA-AD_10x_meta_mini.csv", row.names = 1) %>% 
  mutate(method = "10x")
counts_10x <- read.csv("SEA-AD_10x_counts_mini.csv", row.names = 1)

# filter counts matrices
## Get the common gene names between the two datasets
common_genes <- intersect(colnames(counts_smart), colnames(counts_10x))

## Filter each counts matrix to include only the common genes
counts_smart <- counts_smart[, common_genes]
counts_10x <- counts_10x[, common_genes]

# create Seurat objects
Seu_smart_for_int <- CreateSeuratObject(counts = t(counts_smart), 
                                meta.data = meta_smart) 

Seu_10x_for_int <- CreateSeuratObject(counts = t(counts_10x), 
                                meta.data = meta_10x) 

#rm(counts_smart, counts_10x, meta_smart, meta_10x)
```

## Integration
Choose to either go through the standard preprocessing workflow or the SCTransform workflow.
### Standard method
```{r}
Seu.list <- c(Seu_smart_for_int, Seu_10x_for_int)

#rm(Seu_smart, Seu_10x)

# normalize and identify variable features for each dataset independently
Seu.list <- lapply(X = Seu.list, FUN = function(x) {
  x <- NormalizeData(x, normalization.method = "LogNormalize", scale.factor = 1000000)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 3000)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = Seu.list)

# controlling for differences between within-dataset variabilities (this step is slow)
anchors <- FindIntegrationAnchors(object.list = Seu.list, anchor.features = features)

# this command creates an 'integrated' data assay
Seu_intd_obj <- IntegrateData(anchorset = anchors)

#rm(Seu_smart_for_int, Seu_10x_for_int)
```

### Using SCTransform method
Alternative to above, uses modified normalization method based on regularized negative binomial regression 
```{r}
Seu.list <- c(Seu_smart_for_int, Seu_10x_for_int)

Seu.list <- lapply(X = Seu.list, FUN = function(x) {
  x <- SCTransform(x, variable.features.n = 3000, conserve.memory = T)
})

# select features that are repeatedly variable across datasets for integration
features <- SelectIntegrationFeatures(object.list = Seu.list, nfeatures=3000)

Seu.list <- PrepSCTIntegration(object.list = Seu.list, anchor.features = features)

# controlling for differences between within-dataset variabilities (this step is slow)
anchors <- FindIntegrationAnchors(object.list = Seu.list, normalization.method = "SCT",
                                  anchor.features = features)

# this command creates an 'integrated' data assay
Seu_intd_obj <- IntegrateData(anchorset = anchors, normalization.method = "SCT")

#rm(Seu_smart_for_int, Seu_10x_for_int)
```

## Initial analysis
```{r}
# specify that we will perform downstream analysis on the corrected data, note that 
# the original unmodified data still resides in the 'RNA' assay
DefaultAssay(Seu_intd_obj) <- "integrated"

# Run the standard workflow for visualization and clustering
Seu_intd_obj <- ScaleData(Seu_intd_obj, verbose = FALSE)
```

```{r}
Seu_intd_obj <- RunPCA(Seu_intd_obj, npcs = 30, verbose = FALSE)
ElbowPlot(Seu_intd_obj, ndims = 30)
```

```{r}
Seu_intd_obj <- FindNeighbors(Seu_intd_obj, reduction = "pca", dims = 1:30) %>% 
  FindClusters(resolution = 0.5) %>% 
  RunUMAP(reduction = "pca", dims = 1:30)

# Check out how many cells placed into each cluster
table(Seu_intd_obj$seurat_clusters)
```

## Visualizations
### Examining our clusters
```{r}
DimPlot(Seu_intd_obj, reduction="umap", label=TRUE)
```

### How well did our datasets integrate?
```{r}
# how well did our datasets integrate?
p1 <- DimPlot(Seu_intd_obj, reduction = "umap", group.by = "method")
p2 <- DimPlot(Seu_intd_obj, reduction = "umap", group.by = "class_label")
p1 + p2
```

```{r}
DimPlot(Seu_intd_obj, reduction = "umap", group.by = "subclass_label", 
        split.by = "method", label=TRUE)
```

Looks like our two datasets have different naming conventions for the same cell classes and subclasses. Let's fix this by taking everything in the subclass_label column and making it uppercase:
```{r}
Seu_intd_obj@meta.data$subclass_label <- toupper(Seu_intd_obj@meta.data$subclass_label)
```

### Create a plot to highlight a cell type
```{r}
DimPlot(Seu_intd_obj, reduction = "umap", group.by = "subclass_label",
        cells.highlight = list(VIP=Seu_intd_obj@meta.data %>% 
          filter(subclass_label == "VIP") %>% 
                   row.names(.),
          SST=Seu_intd_obj@meta.data %>% 
          filter(subclass_label =="SST") %>% 
                   row.names(.)),
        cols.highlight = c("slateblue", "cadetblue"))
```

### Create a plot to compare gene expression within clusters between methods
```{r}
FeaturePlot(Seu_intd_obj, features = c("SST", "VIP", "PVALB"), split.by="method")
```

## Find conserved markers
Can use these results to annotate clusters as specific cell types (look with FeaturePlot, rename with RenameIdents)

```{r}
# For performing differential expression after integration, we switch back to the 
# original data
DefaultAssay(Seu_intd_obj) <- "RNA"
```

*Optional:* This next step will take a long time, would recommend using parallelization here. Parallel processing should only be attempted if you think your computer can handle it—if not, skip this chunk and proceed.
```{r}
#install.packages("future")
library(future)
availableCores() # check how many "workers" you can assign
plan("multiprocess", workers = 4) # changes from sequential to parallel
```

Let's look at cluster 9. Based on our earlier figures, cells in this group appear to be astrocytes.
```{r}
Idents(Seu_intd_obj) <- "seurat_clusters"
cluster_9_markers <- FindConservedMarkers(Seu_intd_obj, ident.1 = 9, 
                                          grouping.var = "method",
                                          logfc.threshold = log(2), 
                                          min.pct = 0.25)

# top conserved cluster markers
cluster_9_markers %>% 
  head(n = 10)
```

## Differential expression
Recall that when using "integrated" values, fold change values cannot be directly interpreted as integrated values have become unitless
```{r}
DefaultAssay(Seu_intd_obj) <- "integrated"
# to compare a cluster between methods, must make metadata column combining these 
# traits and set as active identity

Seu_intd_obj$cluster_method <- paste(Idents(Seu_intd_obj), Seu_intd_obj$method, 
                                     sep = "_")

Idents(Seu_intd_obj) <- "cluster_method"

cluster_9_method_de <- FindMarkers(Seu_intd_obj, ident.1 = "9_10x", 
                                   ident.2 = "9_smart", 
                                   logfc.threshold = log(2), min.pct = 0.25)

# top DEG where 10x expresses more
c9_in_10x <- cluster_9_method_de %>% 
  filter(p_val_adj < 0.05) %>% 
  filter(pct.1 > 0.1) %>% 
  arrange(desc(avg_log2FC)) %>% 
  head(n=5) %>% 
  row.names()

# top DEG where smart-seq expresses more
c9_in_smart <- cluster_9_method_de %>% 
  filter(p_val_adj < 0.05) %>% 
  filter(pct.2 > 0.1) %>% 
  arrange(avg_log2FC) %>% 
  head(n=5) %>% 
  row.names()
```

### Before vs. after integration
```{r}
Idents(Seu_intd_obj) <- "seurat_clusters"
DefaultAssay(Seu_intd_obj) <- "RNA"
DotPlot(Seu_intd_obj, features = c(c9_in_10x, c9_in_smart), idents=9, 
        group.by = "method") + RotatedAxis()

DefaultAssay(Seu_intd_obj) <- "integrated"
DotPlot(Seu_intd_obj, features = c(c9_in_10x, c9_in_smart), idents=9, 
        group.by = "method") + RotatedAxis()
```

## Cell type annotation
Let's use our 10x dataset to annotate our Smart-seq dataset

### Using TransferData
In data transfer:
- Seurat does not correct or modify the query expression data.
- Seurat has an option (set by default) to project the PCA structure of a reference onto the query

For the chunk below, use your pre-processed Seurat objects from the last two days.
```{r}
# We want to be able to check how accurate the mapping was, so let's make that
# subclass label fix like we did before:
Seu_10x@meta.data$subclass_label <- toupper(Seu_10x@meta.data$subclass_label)
Seu_smart@meta.data$subclass_label <- toupper(Seu_smart@meta.data$subclass_label)

anchors <- FindTransferAnchors(reference = Seu_10x, query = Seu_smart,
    dims = 1:30, reference.reduction = "pca")
predictions <- TransferData(anchorset = anchors, refdata = Seu_10x$subclass_label,
    dims = 1:30)
smart_query <- AddMetaData(Seu_smart, metadata = predictions)
```

After finding anchors, we use the TransferData() function to classify the query cells based on reference data (a vector of reference cell type labels). TransferData() returns a matrix with predicted IDs and prediction scores, which we can add to the query metadata.

#### How accurate was the annotation transfer?
Note that even after using toupper, there are some subclass_label mismatches, and not all subclasses are present in both datasets... can chose to fix this manually.
```{r}
smart_query$prediction.match <- smart_query$predicted.id == smart_query$subclass_label

table(smart_query$prediction.match)

table(smart_query$predicted.id)
```
After using the 10x as a reference above, you can choose to try using the Smart-seq as a reference to annotate the Smart-seq cells and see how this performs.

### Using MapQuery
An alternative to the above, based on projecting query dataset onto the UMAP structure of the reference dataset.

```{r}
anchors <- FindTransferAnchors(reference = Seu_10x, query = Seu_smart,
    dims = 1:30, reference.reduction = "pca")

Seu_10x <- RunUMAP(Seu_10x, dims = 1:30, return.model = TRUE)
Seu_smart <- RunUMAP(Seu_smart, dims = 1:40, return.model = TRUE)

smart_query <- MapQuery(anchorset = anchors, reference = Seu_10x, query = Seu_smart,
    refdata = list(subclass_label = "subclass_label"), reference.reduction = "pca", 
    reduction.model = "umap")
```

#### Let's see how well that worked:
```{r}
p1 <- DimPlot(Seu_10x, reduction = "umap", group.by = "subclass_label", label = TRUE, label.size = 3,
    repel = TRUE) + NoLegend() + ggtitle("Reference annotations")
p2 <- DimPlot(smart_query, reduction = "ref.umap", group.by = "predicted.subclass_label", label = TRUE,
    label.size = 3, repel = TRUE) + NoLegend() + ggtitle("Query transferred labels")
p1 + p2
```
